---
title: "repro-research-1"
author: "Jason Chui"
date: "1 August 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible Research Peer 1
## Getting and pre-processing data
Read in the data in activity.csv into the "df" variable
```{r}
df = read.csv("activity.csv")
```

Transform the date column type from factor into Date
```{r}
df$date = as.Date(df$date, format="%Y-%m-%d")
```

## What is the total steps taken by day?
To find the total steps by day, we can use tapply on df:
```{r}
totalSteps <- tapply(df$steps, df$date, FUN=sum, na.rm=TRUE)
totalSteps
```

Another way to group the steps by day is to use aggregate:
```{r}
aggregate(steps ~ date, data = df, sum, na.rm = TRUE)
```

To plot a histogram of totalSteps, we are going to use ggplot2. We need to hence load ggplot2
```{r}
library(ggplot2)
```

We shall now use qplot to plot the histogram:
```{r}
qplot(totalSteps, binwidth=1000, xlab="Total steps each day", ylab="Frequency")
```

The mean and median of the total steps by day can be calculated as such:
```{r}
mean(totalSteps)
median(totalSteps)
```

## What is average steps per 5-minute interval?
To find out the average steps per 5-minute interval, we shall first aggregate the steps by the interval, and apply the mean() function to the aggregate.
```{r}
averageSteps <- aggregate(x=list(steps=df$steps), by=list(interval=df$interval),
                      FUN=mean, na.rm=TRUE)
```

We can use ggplot to plot this line series.
```{r}
ggplot(data=averageSteps, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken")
```

```{r}
averageSteps[which.max(averageSteps$steps),]
```